---
title: "ANCOM-BC"
author: 
  - Huang Lin$^1$
  - $^1$Department of Biostatistics, University of Pittsburgh, 130 De Soto Street, Pittsburgh, PA 15261 USA
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: 
    toc: true
    theme: united
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{ANCOMBC}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA, 
                      fig.width = 6.25, fig.height = 5)

library(tidyverse)
library(microbiome)
library(magrittr)
library(qwraps2)
library(ANCOMBC)
library(DT)
options(DT.options = list(
  initComplete = JS("function(settings, json) {",
  "$(this.api().table().header()).css({'background-color': 
  '#000', 'color': '#fff'});","}")))
```

# 1. Introduction

Analysis of Compositions of Microbiomes with Bias Correction (ANCOM-BC) [@lin2020analysis] 
is a methodology of differential abundance (DA) analysis for microbial absolute
abundance data. ANCOM-BC estimates the unknown sampling fractions, corrects 
the bias induced by their differences among samples, and identifies taxa that 
are differentially abundant according to the covariate of interest.

For more details, please refer to the 
[ANCOM-BC](https://doi.org/10.1038/s41467-020-17041-7) paper.

# 2. Installation

Download package. 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ANCOMBC")
```

Load the package. 

```{r load, eval=FALSE}
library(ANCOMBC)
```

# 3. Running ANCOMBC

## 3.1 Intestinal microbiota profiling data

The HITChip Atlas data set [@lahti2014tipping] is available via the 
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq] 
format, and via Data Dryad in tabular format. 
This data set comes with 130 genus-like taxonomic groups across 1006 western 
adults with no reported health complications. 
Some subjects have also short time series.

```{r importData}
data(atlas1006) 
# Subset to baseline
pseq = subset_samples(atlas1006, time == 0)
# Re-code the bmi group
sample_data(pseq)$bmi_group = recode(sample_data(pseq)$bmi_group,
                                     underweight = "lean",
                                     lean = "lean",
                                     overweight = "overweight",
                                     obese = "obese",
                                     severeobese = "obese",
                                     morbidobese = "obese")
# Re-code the nationality group
sample_data(pseq)$nation = recode(sample_data(pseq)$nationality,
                                  Scandinavia = "NE",
                                  UKIE = "NE",
                                  SouthEurope = "SE",
                                  CentralEurope = "CE",
                                  EasternEurope = "EE")

# Aggregate to phylum level
phylum_data = aggregate_taxa(pseq, "Phylum")
```

## 3.2 Data summary

```{r dataSummary, results = "asis"}
options(qwraps2_markup = "markdown")
summary_template =
  list("Age" =
       list("min" = ~ min(.data$age, na.rm = TRUE),
            "max" = ~ max(.data$age, na.rm = TRUE),
            "mean (sd)" = ~ mean_sd(.data$age, na_rm = TRUE, 
                                    show_n = "never")),
       "Gender" =
       list("F" = ~ n_perc0(.data$sex == "female", na_rm = TRUE),
            "M" = ~ n_perc0(.data$sex == "male", na_rm = TRUE),
            "NA" = ~ n_perc0(is.na(.data$sex))),
       "Nationality" =
       list("Central Europe" = ~ n_perc0(.data$nation == "CE", na_rm = TRUE),
            "Eastern Europe" = ~ n_perc0(.data$nation == "EE", na_rm = TRUE),
            "Northern Europe" = ~ n_perc0(.data$nation == "NE", na_rm = TRUE),
            "Southern Europe" = ~ n_perc0(.data$nation == "SE", na_rm = TRUE),
            "US" = ~ n_perc0(.data$nation == "US", na_rm = TRUE),
            "NA" = ~ n_perc0(is.na(.data$nation))),
       "BMI" =
       list("Lean" = ~ n_perc0(.data$bmi_group == "lean", na_rm = TRUE),
            "Overweight" = ~ n_perc0(.data$bmi_group == "overweight", 
                                     na_rm = TRUE),
            "Obese" = ~ n_perc0(.data$bmi_group == "obese", na_rm = TRUE),
            "NA" = ~ n_perc0(is.na(.data$bmi_group)))
       )
data_summary = summary_table(meta(pseq), summary_template)
data_summary
```

1. The number of samples: `r nsamples(phylum_data)`.

2. The number of phyla: `r ntaxa(phylum_data)`.

## 3.3 Running ancombc function

```{r ancombc}
out = ancombc(phyloseq = phylum_data, formula = "age + nation + bmi_group", 
              p_adj_method = "holm", zero_cut = 0.90, lib_cut = 1000, 
              group = "nation", struc_zero = TRUE, neg_lb = TRUE, tol = 1e-5, 
              max_iter = 100, conserve = TRUE, alpha = 0.05, global = TRUE)

res = out$res
res_global = out$res_global
```

## 3.4 ANCOMBC primary result

Result from the ANCOM-BC log-linear model to determine taxa that are 
differentially abundant according to the covariate of interest. It contains: 
1) coefficients; 2) standard errors; 3) test statistics; 4) p-values; 
5) adjusted p-values; 6) indicators whether the taxon is differentially 
abundant (TRUE) or not (FALSE).

### 3.41 Coefficients

```{r}
tab_coef = res$beta
col_name = c("Age", "EE - CE", "NE - CE", "SE - CE", "US - CE", 
             "Oerweight - Lean", "Obese - Lean")
colnames(tab_coef) = col_name
tab_coef %>% datatable(caption = "Coefficients from the Primary Result") %>%
      formatRound(col_name, digits = 2)
```

### 3.42 SEs

```{r}
tab_se = res$se
colnames(tab_se) = col_name
tab_se %>% datatable(caption = "SEs from the Primary Result") %>%
      formatRound(col_name, digits = 2)
```

### 3.43 Test statistics

```{r}
tab_w = res$W
colnames(tab_w) = col_name
tab_w %>% datatable(caption = "Test Statistics from the Primary Result") %>%
      formatRound(col_name, digits = 2)
```

### 3.44 P-values

```{r}
tab_p = res$p_val
colnames(tab_p) = col_name
tab_p %>% datatable(caption = "P-values from the Primary Result") %>%
      formatRound(col_name, digits = 2)
```

### 3.45 Adjusted p-values

```{r}
tab_q = res$q
colnames(tab_q) = col_name
tab_q %>% datatable(caption = "Adjusted p-values from the Primary Result") %>%
      formatRound(col_name, digits = 2)
```

### 3.46 Differentially abundant taxa

```{r}
tab_diff = res$diff_abn
colnames(tab_diff) = col_name
tab_diff %>% 
  datatable(caption = "Differentially Abundant Taxa 
            from the Primary Result")
```

### 3.45 Visualizations for "age"

*"Age" is a continuous variable.*

```{r}
df_fig1 = data.frame(res$beta * res$diff_abn, check.names = FALSE) %>% 
  rownames_to_column("taxon_id")
df_fig2 = data.frame(res$se * res$diff_abn, check.names = FALSE) %>% 
  rownames_to_column("taxon_id")
colnames(df_fig2)[-1] = paste0(colnames(df_fig2)[-1], "SD")
df_fig = df_fig1 %>% left_join(df_fig2, by = "taxon_id") %>%
  transmute(taxon_id, age, ageSD) %>%
  filter(age != 0) %>% arrange(desc(age)) %>%
  mutate(group = ifelse(age > 0, "g1", "g2"))
df_fig$taxon_id = factor(df_fig$taxon_id, levels = df_fig$taxon_id)
  
p = ggplot(data = df_fig, 
           aes(x = taxon_id, y = age, fill = group, color = group)) + 
  geom_bar(stat = "identity", width = 0.7, 
           position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = age - ageSD, ymax = age + ageSD), width = 0.2,
                position = position_dodge(0.05), color = "black") + 
  labs(x = NULL, y = "Log fold change", 
       title = "Waterfall Plot for the Age Effect") + 
  theme_bw() + 
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1))
p
```

## 3.5 ANCOMBC global test result

Result from the ANCOM-BC global test to determine taxa that are 
differentially abundant between three or more groups of multiple samples.
It contains: 1) test statistics; 2) p-values; 3) adjusted p-values; 
4) indicators whether the taxon is differentially abundant (TRUE) or not 
(FALSE).

### 3.51 Test statistics

```{r}
tab_w = res_global[, "W", drop = FALSE]
colnames(tab_w) = "Nationality"
tab_w %>% datatable(caption = "Test Statistics 
                    from the Global Test Result") %>%
      formatRound(c("Nationality"), digits = 2)
```

### 3.52 P-values

```{r}
tab_p = res_global[, "p_val", drop = FALSE]
colnames(tab_p) = "Nationality"
tab_p %>% datatable(caption = "P-values 
                    from the Global Test Result") %>%
      formatRound(c("Nationality"), digits = 2)
```

### 3.53 Adjusted p-values

```{r}
tab_q = res_global[, "q_val", drop = FALSE]
colnames(tab_q) = "Nationality"
tab_q %>% datatable(caption = "Adjusted p-values 
                    from the Global Test Result") %>%
      formatRound(c("Nationality"), digits = 2)
```

### 3.54 Differentially abundant taxa

```{r}
tab_diff = res_global[, "diff_abn", drop = FALSE]
colnames(tab_diff) = "Nationality"
tab_diff %>% datatable(caption = "Differentially Abundant Taxa 
                       from the Global Test Result")
```

# Session information

```{r sessionInfo, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```

# References









