---
title: "ANCOM-BC2 Tutorial"
author: 
  - Huang Lin$^1$
  - $^1$University of Maryland, College Park, MD 20742
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{ANCOM-BC2 Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, message = FALSE, warning = FALSE, comment = NA}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA, 
                      fig.width = 6.25, fig.height = 5)
library(ANCOMBC)
library(tidyverse)
library(DT)
options(DT.options = list(
  initComplete = JS("function(settings, json) {",
  "$(this.api().table().header()).css({'background-color': 
  '#000', 'color': '#fff'});","}")))

# It appears to be a package compatibility issue between the release version of 
# phyloseq and lme4, a fresh installation of phyloseq might be needed
# See this post: https://github.com/lme4/lme4/issues/743
# remotes::install_github("joey711/phyloseq", force = TRUE)
```

# 1. Introduction

Analysis of Compositions of Microbiomes with Bias Correction 2 (ANCOM-BC2) is
a methodology for performing differential abundance (DA) analysis of microbiome
count data. This version extends and refines the previously published
Analysis of Compositions of Microbiomes with Bias Correction (ANCOM-BC)
methodology [@lin2020analysis] in several ways as follows:

1. **Bias correction**: ANCOM-BC2 estimates and corrects both the
sample-specific (sampling fraction) as well as
taxon-specific (sequencing efficiency) biases.

2. **Regularization of variance**: Inspired by Significance Analysis of
Microarrays (SAM) [@tusher2001significance] methodology, a small positive
constant is added to the denominator of ANCOM-BC2 test statistic corresponding
to each taxon to avoid the significance due to extremely small standard errors,
especially for rare taxa. By default, we used the 5-th percentile of the
distribution of standard errors for each fixed effect as the regularization
factor.

3. **Sensitivity analysis for the pseudo-count addition**: Like other
differential abundance analysis methods, ANCOM-BC2 applies a log transformation
to the observed counts, and the zero counts require the addition of a
pseudo-count. The choice of pseudo-count affects the results and can inflate
the false positive rate [@costea2014fair; @paulson2014reply]. ANCOM-BC2
therefore repeats the analysis over a set of pseudo-count values and reports,
for each taxon, whether the conclusion about its significance is unchanged. The
stage at which the pseudo-counts are added is set by the `conservative`
argument, and the two decision rules are stated in Section 7.4. Both rules
compare significance at the level alpha, so the results of the sensitivity
analysis change when alpha changes.

4. **Multi-group comparisons and repeated measurements**: The ANCOM-BC2
methodology extends ANCOM-BC for multiple groups and repeated measurements
as follows:

    + Multiple pairwise comparisons: When performing multiple pairwise
    comparisons, the mixed directional false discovery rate (mdFDR) should be
    taken into account. The mdFDR is the combination of false
    discovery rate due to multiple testing, multiple pairwise comparisons,
    and directional tests within each pairwise comparison. For example, suppose
    we have five taxa and three experimental groups: g1, g2, and g3.
    Thus, we are performing five tests corresponding to five taxa.
    For each taxon, we are also conducting three pairwise comparisons
    (g1 vs. g2, g2 vs. g3, and g1 vs. g3). Within each pairwise comparison,
    we wish to determine if the abundance has increased or decreased or did not
    change (direction of the effect size). Errors could occur in each step.
    The overall false discovery rate is controlled by the mdFDR methodology we
    adopted from [@guo2010controlling; @grandhi2016multiple].

    + Multiple pairwise comparisons against a pre-specified group
    (e.g., Dunnett's type of test): We use the same set-up as in the multiple
    pairwise comparisons but use the Dunnett-type modification described
    in [@grandhi2016multiple] to control the mdFDR which is more powerful.

    + Pattern analysis for ordered groups: In some instances, researchers are
    interested in discovering abundance patterns of each taxon over the ordered
    groups, for example, groups based on the health condition of subjects
    (e.g., lean, overweight, obese). In such cases, in addition to pairwise
    comparison, one may be interested in identifying taxa whose abundances are
    increasing or decreasing or have other patterns over the groups.
    We adopted methodologies from [@jelsema2016clme] to
    perform pattern analysis under the ANCOM-BC2 framework.

**A clarification regarding Structural zeros**: A taxon is considered to have
structural zeros in some (>=1) groups if it is completely (or nearly completely)
missing in these groups. For instance, suppose there are three groups:
g1, g2, and g3. If the counts of taxon A in g1 are 0, but they are nonzero
in g2 and g3, then taxon A will be considered to contain structural zeros in g1.
In this example, taxon A is declared to be differentially abundant between
g1 and g2, g1 and g3, and consequently, it is globally differentially abundant
with respect to this group variable. Such taxa are not further analyzed using
ANCOM-BC2, but the results are summarized in a separate table.

# 2. Installation

Download the package.

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ANCOMBC")
```

Load the package. 

```{r load, eval=FALSE}
library(ANCOMBC)
```

# 3. Input formats and a minimal call

The `ancombc2` function accepts three input formats.

1. A `phyloseq` object, supplied as `data`. Aggregation to a taxonomic level is
requested with `tax_level`.

2. A `TreeSummarizedExperiment` object, supplied as `data` together with
`assay_name`. Aggregation is requested with `tax_level`.

3. A count `matrix` or `data.frame`, supplied as `data` together with
`meta_data`, and with `aggregate_data` when the analysis is run on aggregated
counts. Taxa are rows and samples are columns; the row names of `data` are the
taxon names, the column names of `data` are the sample names, and the row names
of `meta_data` are the same sample names. Aggregation is not performed for this
input format, so the counts are aggregated to the taxonomic level of interest
beforehand.

A minimal call specifies the data, the taxonomic level, the fixed-effects
formula, and the group variable. All remaining arguments take their default
values.

```{r, eval=FALSE}
output = ancombc2(data = pseq, tax_level = "Family",
                  fix_formula = "age + region + bmi", group = "bmi")
```

# 4. Run ANCOM-BC2 on a real cross-sectional dataset

## 4.1 Import example data

The HITChip Atlas dataset contains genus-level microbiota profiling with
HITChip for 1006 western adults with no reported health complications,
reported in [@lahti2014tipping]. The dataset is available via the
`microbiome` R package [@lahti2017tools] in `phyloseq` [@mcmurdie2013phyloseq]
format. In this tutorial, we consider the following covariates:

* Continuous covariates: "age"

* Categorical covariates: "region", "bmi"

* The group variable of interest: "bmi"

    + Three groups: "lean", "overweight", "obese"

    + The reference group: "obese"

```{r, child = "children/child_atlas1006.Rmd"}
```

## 4.2 Run `ancombc2` function using the `phyloseq` object

The p-values are adjusted by the Holm-Bonferroni method rather than by the
Benjamini-Hochberg procedure. The Holm-Bonferroni method is valid under any
dependence structure among the p-values and is robust to inaccuracies in them,
which are common in DA analysis. The figures below display only the taxa that
remain significant after the Holm-Bonferroni adjustment.

```{r}
set.seed(123)
# It should be noted that we have set the number of bootstrap samples (B) equal 
# to 10 in the 'trend_control' function for computational expediency. 
# However, it is recommended that users utilize the default value of B, 
# which is 100, or larger values for optimal performance.
output = ancombc2(data = pseq, tax_level = "Family",
                  fix_formula = "age + region + bmi", rand_formula = NULL,
                  p_adj_method = "holm", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "bmi", struc_zero = TRUE, neg_lb = TRUE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
                  iter_control = list(tol = 1e-2, max_iter = 20, 
                                      verbose = TRUE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
                  trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
                                                              nrow = 2, 
                                                              byrow = TRUE),
                                                       matrix(c(-1, 0, 1, -1),
                                                              nrow = 2, 
                                                              byrow = TRUE),
                                                       matrix(c(1, 0, 1, -1),
                                                              nrow = 2, 
                                                              byrow = TRUE)),
                                       node = list(2, 2, 1),
                                       B = 10))
```

Tests for interactions are supported by specifying the interactions in
`fix_formula`. Interactions between variables must be written with the `*`
operator, as the `:` operator is not recognized by the `ancombc2` function and
results in an error.

When the `group` variable is involved in an interaction term, only its main
effect is used in multi-group comparisons.

```{r}
set.seed(123)
output_interact = ancombc2(data = pseq, tax_level = "Family",
                          fix_formula = "age * bmi + region", rand_formula = NULL,
                          p_adj_method = "holm", pseudo_sens = TRUE,
                          prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                          group = "bmi", struc_zero = TRUE, neg_lb = TRUE,
                          alpha = 0.05, n_cl = 2, verbose = TRUE)

head(output_interact$res) %>%
    dplyr::select(starts_with("lfc_age")) %>%
    mutate_if(is.numeric, function(x) round(x, 2)) %>%
    datatable(caption = "Example of including interaction terms")
```

## 4.3 How to read the output

The primary result `output$res` is a data frame with one row per taxon. Its
columns are grouped by prefix, and the suffix of a column name is the fixed
effect it refers to, for example `age` or `bmilean`. Let $\alpha$ denote the
significance level, set by the `alpha` argument.

* `lfc_`: the estimated log fold change (lfc), on the natural log scale.

* `se_`: the standard error of the log fold change. The variance is regularized
before the square root is taken, by adding the `s0_perc` quantile of the
variance distribution of the corresponding fixed effect.

* `W_`: the test statistic, $W = \text{lfc}/\text{se}$.

* `p_`: the two-sided p-value of the hypothesis $\text{lfc} = 0$.

* `q_`: the p-value adjusted across taxa by `p_adj_method`.

* `diff_`: the indicator of differential abundance, $\text{diff\_abn} =
\mathbf{1}\{q \le \alpha\}$.

* `passed_ss_`: the indicator that the taxon is insensitive to the pseudo-count
addition. Under the default `conservative = TRUE`, with $s$ the sensitivity
score of the taxon,
$$ \text{passed\_ss} = \mathbf{1}\{s = 0 \text{ or } s = 1\}. $$

* `diff_robust_`: the conjunction of the two preceding indicators,
$$ \text{diff\_robust} = \text{diff\_abn} \wedge \text{passed\_ss}. $$

A taxon that is nonsignificant in every run of the sensitivity analysis also has
$s = 1$ and therefore passes, so `passed_ss` alone is not an indicator of
significance. The sensitivity scores themselves are returned in `output$ss_tab`,
with one column per fixed effect and per test. Section 7.4 states the
sensitivity score and the decision rule for both settings of `conservative`.

The results of the global test (`output$res_global`), the pairwise directional
test (`output$res_pair`), the Dunnett's type of test (`output$res_dunn`), and
the pattern analysis (`output$res_trend`) use the same column groups. The
global test and the pattern analysis report one test per taxon, so their
indicator columns are `diff_abn`, `passed_ss`, and `diff_robust_abn`.

## 4.4 Structural zeros (taxon presence/absence)

```{r}
tab_zero = output$zero_ind
tab_zero %>%
    datatable(caption = "The detection of structural zeros")
```

## 4.5 ANCOM-BC2 primary analysis {.tabset}

The primary output identifies the taxa that are differentially abundant with
respect to each fixed effect. The column groups are defined in Section 4.3.

```{r}
res_prim = output$res
```

### Results for age

In the waterfall plot below, each bar is a log fold change, on the natural log
scale. The taxon labels in green mark the taxa whose `diff_robust` indicator is
TRUE, that is, the taxa that are differentially abundant and insensitive to the
pseudo-count addition.

```{r}
df_age = res_prim %>%
    dplyr::select(taxon, ends_with("age")) 
df_fig_age = df_age %>%
    dplyr::filter(diff_age == 1) %>% 
    dplyr::arrange(desc(lfc_age)) %>%
    dplyr::mutate(direct = ifelse(lfc_age > 0, "Positive LFC", "Negative LFC"),
                  color = ifelse(diff_robust_age, "aquamarine3", "black"))
df_fig_age$taxon = factor(df_fig_age$taxon, levels = df_fig_age$taxon)
df_fig_age$direct = factor(df_fig_age$direct, 
                           levels = c("Positive LFC", "Negative LFC"))
  
fig_age = df_fig_age %>%
    ggplot(aes(x = taxon, y = lfc_age, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = lfc_age - se_age, ymax = lfc_age + se_age), 
                  width = 0.2, position = position_dodge(0.05), color = "black") + 
    labs(x = NULL, y = "Log fold change", 
         title = "Log fold changes as one unit increase of age") + 
    scale_fill_discrete(name = NULL) +
    scale_color_discrete(name = NULL) +
    theme_bw() + 
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.minor.y = element_blank(),
          axis.text.x = element_text(angle = 60, hjust = 1,
                                     color = df_fig_age$color))
fig_age
```

### Results for bmi

In the heatmap below, each cell is a log fold change, on the natural log scale.
The entries in green mark the comparisons whose `diff_robust` indicator is TRUE.

```{r}
df_bmi = res_prim %>%
    dplyr::select(taxon, contains("bmi")) 
df_fig_bmi1 = df_bmi %>%
    dplyr::filter(diff_bmilean == 1 | 
                    diff_bmioverweight == 1) %>%
    dplyr::mutate(lfc1 = ifelse(diff_bmioverweight == 1, 
                                round(lfc_bmioverweight, 2), 0),
                  lfc2 = ifelse(diff_bmilean == 1, 
                                round(lfc_bmilean, 2), 0)) %>%
    tidyr::pivot_longer(cols = lfc1:lfc2, 
                        names_to = "group", values_to = "value") %>%
    dplyr::arrange(taxon)

df_fig_bmi2 = df_bmi %>%
    dplyr::filter(diff_bmilean == 1 | 
                    diff_bmioverweight == 1) %>%
    dplyr::mutate(lfc1 = ifelse(diff_robust_bmioverweight, 
                                "aquamarine3", "black"),
                  lfc2 = ifelse(diff_robust_bmilean, 
                                "aquamarine3", "black")) %>%
    tidyr::pivot_longer(cols = lfc1:lfc2, 
                        names_to = "group", values_to = "color") %>%
    dplyr::arrange(taxon)

df_fig_bmi = df_fig_bmi1 %>%
    dplyr::left_join(df_fig_bmi2, by = c("taxon", "group"))

df_fig_bmi$group = recode(df_fig_bmi$group, 
                          `lfc1` = "Overweight - Obese",
                          `lfc2` = "Lean - Obese")
df_fig_bmi$group = factor(df_fig_bmi$group, 
                          levels = c("Overweight - Obese",
                                     "Lean - Obese"))
  
lo = floor(min(df_fig_bmi$value))
up = ceiling(max(df_fig_bmi$value))
mid = (lo + up)/2
fig_bmi = df_fig_bmi %>%
  ggplot(aes(x = group, y = taxon, fill = value)) + 
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "white", midpoint = mid, limit = c(lo, up),
                       name = NULL) +
  geom_text(aes(group, taxon, label = value, color = color), size = 4) +
  scale_color_identity(guide = "none") +
  labs(x = NULL, y = NULL, title = "Log fold changes as compared to obese subjects") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
fig_bmi
```

## 4.6 ANCOM-BC2 global test

The ANCOM-BC2 global test identifies taxa that are differentially abundant
between at least two groups when three or more groups are compared. Here the
groups are the "lean", "overweight", and "obese" categories. The column groups
are defined in Section 4.3.

In the heatmap below, each cell is a log fold change, on the natural log scale.
The taxa in green are those whose `diff_robust_abn` indicator is TRUE.

```{r}
res_global = output$res_global
df_bmi = res_prim %>%
    dplyr::select(taxon, contains("bmi")) 
df_fig_global = df_bmi %>%
    dplyr::left_join(res_global %>%
                       dplyr::transmute(taxon, 
                                        diff_bmi = diff_abn, 
                                        diff_robust_bmi = diff_robust_abn)) %>%
    dplyr::filter(diff_bmi == 1) %>%
    dplyr::mutate(lfc_overweight = lfc_bmioverweight,
                  lfc_lean = lfc_bmilean,
                  color = ifelse(diff_robust_bmi, "aquamarine3", "black")) %>%
    dplyr::transmute(taxon,
                     `Overweight - Obese` = round(lfc_overweight, 2),
                     `Lean - Obese` = round(lfc_lean, 2), 
                     color = color) %>%
    tidyr::pivot_longer(cols = `Overweight - Obese`:`Lean - Obese`, 
                        names_to = "group", values_to = "value") %>%
    dplyr::arrange(taxon)

df_fig_global$group = factor(df_fig_global$group, 
                             levels = c("Overweight - Obese",
                                        "Lean - Obese"))
  
lo = floor(min(df_fig_global$value))
up = ceiling(max(df_fig_global$value))
mid = (lo + up)/2
fig_global = df_fig_global %>%
  ggplot(aes(x = group, y = taxon, fill = value)) + 
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "white", midpoint = mid, limit = c(lo, up),
                       name = NULL) +
  geom_text(aes(group, taxon, label = value), color = "black", size = 4) +
  labs(x = NULL, y = NULL, title = "Log fold changes for globally significant taxa") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.y = element_text(color = df_fig_global %>%
                                       dplyr::distinct(taxon, color) %>%
                                       .$color))
fig_global
```

## 4.7 ANCOM-BC2 multiple pairwise comparisons

The ANCOM-BC2 pairwise directional test identifies taxa that are differentially
abundant between any two groups within a set of three or more groups, while
controlling the mdFDR. The column groups are defined in Section 4.3.

In the heatmap below, each cell is a log fold change, on the natural log scale.
The entries in green mark the comparisons whose `diff_robust` indicator is TRUE.

```{r}
res_pair = output$res_pair

df_fig_pair1 = res_pair %>%
    dplyr::filter(diff_bmioverweight == 1 |
                      diff_bmilean == 1 | 
                      diff_bmilean_bmioverweight == 1) %>%
    dplyr::mutate(lfc1 = ifelse(diff_bmioverweight == 1, 
                                round(lfc_bmioverweight, 2), 0),
                  lfc2 = ifelse(diff_bmilean == 1, 
                                round(lfc_bmilean, 2), 0),
                  lfc3 = ifelse(diff_bmilean_bmioverweight == 1, 
                                round(lfc_bmilean_bmioverweight, 2), 0)) %>%
    tidyr::pivot_longer(cols = lfc1:lfc3, 
                        names_to = "group", values_to = "value") %>%
    dplyr::arrange(taxon)

df_fig_pair2 = res_pair %>%
    dplyr::filter(diff_bmioverweight == 1 |
                      diff_bmilean == 1 | 
                      diff_bmilean_bmioverweight == 1) %>%
    dplyr::mutate(lfc1 = ifelse(diff_robust_bmioverweight, 
                                "aquamarine3", "black"),
                  lfc2 = ifelse(diff_robust_bmilean, 
                                "aquamarine3", "black"),
                  lfc3 = ifelse(diff_robust_bmilean_bmioverweight, 
                                "aquamarine3", "black")) %>%
    tidyr::pivot_longer(cols = lfc1:lfc3, 
                        names_to = "group", values_to = "color") %>%
    dplyr::arrange(taxon)

df_fig_pair = df_fig_pair1 %>%
    dplyr::left_join(df_fig_pair2, by = c("taxon", "group"))

df_fig_pair$group = recode(df_fig_pair$group, 
                          `lfc1` = "Overweight - Obese",
                          `lfc2` = "Lean - Obese",
                          `lfc3` = "Lean - Overweight")
df_fig_pair$group = factor(df_fig_pair$group, 
                          levels = c("Overweight - Obese",
                                     "Lean - Obese", 
                                     "Lean - Overweight"))

lo = floor(min(df_fig_pair$value))
up = ceiling(max(df_fig_pair$value))
mid = (lo + up)/2
fig_pair = df_fig_pair %>%
    ggplot(aes(x = group, y = taxon, fill = value)) + 
    geom_tile(color = "black") +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         na.value = "white", midpoint = mid, limit = c(lo, up),
                         name = NULL) +
    geom_text(aes(group, taxon, label = value, color = color), size = 4) +
    scale_color_identity(guide = FALSE) +
    labs(x = NULL, y = NULL, title = "Log fold changes as compared to obese subjects") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
fig_pair
```

## 4.8 ANCOM-BC2 multiple pairwise comparisons against a pre-specified group (Dunnett's type of test)

The Dunnett's test [@dunnett1955multiple; @dunnett1991step; @dunnett1992step]
compares several experimental groups against a control or reference group. The
ANCOM-BC2 Dunnett's type of test applies this framework and controls the mdFDR.
The primary results control for multiple testing across taxa but not for
multiple comparisons between groups, so, unlike the Dunnett's type of test,
they do not control the mdFDR.

Here the comparisons are "lean" and "overweight" against the reference group
"obese". The column groups are defined in Section 4.3.

In the heatmap below, each cell is a log fold change, on the natural log scale.
The entries in green mark the comparisons whose `diff_robust` indicator is TRUE.

```{r}
res_dunn = output$res_dunn

df_fig_dunn1 = res_dunn %>%
    dplyr::filter(diff_bmioverweight == 1 | 
                      diff_bmilean == 1) %>%
    dplyr::mutate(lfc1 = ifelse(diff_bmioverweight == 1, 
                                round(lfc_bmioverweight, 2), 0),
                  lfc2 = ifelse(diff_bmilean == 1, 
                                round(lfc_bmilean, 2), 0)) %>%
    tidyr::pivot_longer(cols = lfc1:lfc2, 
                        names_to = "group", values_to = "value") %>%
    dplyr::arrange(taxon)

df_fig_dunn2 = res_dunn %>%
    dplyr::filter(diff_bmioverweight == 1 | 
                      diff_bmilean == 1) %>%
    dplyr::mutate(lfc1 = ifelse(diff_robust_bmioverweight, 
                                "aquamarine3", "black"),
                  lfc2 = ifelse(diff_robust_bmilean == 1, 
                                "aquamarine3", "black")) %>%
    tidyr::pivot_longer(cols = lfc1:lfc2, 
                        names_to = "group", values_to = "color") %>%
    dplyr::arrange(taxon)

df_fig_dunn = df_fig_dunn1 %>%
    dplyr::left_join(df_fig_dunn2, by = c("taxon", "group"))

df_fig_dunn$group = recode(df_fig_dunn$group, 
                          `lfc1` = "Overweight - Obese",
                          `lfc2` = "Lean - Obese")
df_fig_dunn$group = factor(df_fig_dunn$group, 
                          levels = c("Overweight - Obese",
                                     "Lean - Obese"))

lo = floor(min(df_fig_dunn$value))
up = ceiling(max(df_fig_dunn$value))
mid = (lo + up)/2
fig_dunn = df_fig_dunn %>%
  ggplot(aes(x = group, y = taxon, fill = value)) + 
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "white", midpoint = mid, limit = c(lo, up),
                       name = NULL) +
  geom_text(aes(group, taxon, label = value, color = color), size = 4) +
    scale_color_identity(guide = FALSE) +
  labs(x = NULL, y = NULL, title = "Log fold changes as compared to obese subjects") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
fig_dunn
```

## 4.9 ANCOM-BC2 pattern analysis {.tabset}

In certain experiments, groups inherently follow an order, such as in
dose-response studies. In these cases, we're interested in discerning if
microbial abundances align with specific patterns. Such patterns might manifest
as monotonically increasing, monotonically decreasing, or an umbrella shape.

ANCOM-BC2 pattern analysis is able to identify potential patterns by testing the
contrast:
$$Ax \ge 0$$
where $A$ is the contrast matrix and $x$ is the vector of parameters.

For instance, in this example, we want to identify taxa that are monotonically
increasing across "obese", "overweight", and "lean". Note that "obese" is the
reference group, and the parameters we can estimate are the differences as
compared to the reference group, i.e.,
$$x = (\text{overweight - obese}, \text{lean - obese})^T$$
To test the monotonically increasing trend:
$$H_0: \text{obese} = \text{overweight} = \text{lean} \\
H_1: \text{obese} \le \text{overweight} \le \text{lean} \quad \text{with at least one strict inequality}$$

We are essentially testing:
$$H_0: 0 = \text{overweight - obese} = \text{lean - obese} \\
H_1: 0 \le \text{overweight - obese} \le \text{lean - obese} \quad \text{with at least one strict inequality}$$

Thus, we shall specify the contrast matrix $A$ as
$$A = \begin{bmatrix} 1 & 0 \\ -1 & 1 \end{bmatrix}$$
The first row of $A$ matrix, $(1, 0)$, indicates that $\text{"overweight - obese"} \ge 0$, and the second row, $(-1, 1)$ represents $\text{"lean - obese"} - \text{"overweight - obese"} \ge 0$.

Similarly, to test for the monotonically decreasing trend:
$$H_0: \text{obese} = \text{overweight} = \text{lean} \\
H_1: \text{obese} \ge \text{overweight} \ge \text{lean} \quad \text{with at least one strict inequality}$$
We shall specify the contrast matrix $A$ as
$$A = \begin{bmatrix} -1 & 0 \\ 1 & -1 \end{bmatrix}$$
Lastly, to test for an umbrella trend:
$$H_0: \text{obese} = \text{overweight} = \text{lean} \\
H_1: \text{obese} \le \text{overweight} \ge \text{lean} \quad \text{with at least one strict inequality}$$
We shall specify the contrast matrix $A$ as
$$A = \begin{bmatrix} 1 & 0 \\ 1 & -1 \end{bmatrix}$$

For testing monotonic trend (increasing or decreasing), one should specify the
`node` parameter in `trend_control` as the last position of `x`.
In this example, the vector of parameters $x$ is of length 2, thus, the last
position is 2. For testing umbrella shape, such as in the above umbrella shape
example, one should set `node` as the position of the turning point of `x`.
In this example, the turning position is `overweight`, thus, `node = 1`.

The monotonically increasing, the monotonically decreasing, and the umbrella
trend are tested in this example. The column groups of the result are defined
in Section 4.3.

In the heatmap below, each cell is a log fold change, on the natural log scale.
The taxa in green are those whose `diff_robust_abn` indicator is TRUE.

```{r}
res_trend = output$res_trend

df_fig_trend = res_trend %>%
    dplyr::filter(diff_abn == 1) %>%
    dplyr::mutate(lfc1 = round(lfc_bmioverweight, 2),
                  lfc2 = round(lfc_bmilean, 2),
                  color = ifelse(diff_robust_abn, "aquamarine3", "black")) %>%
    tidyr::pivot_longer(cols = lfc1:lfc2, 
                        names_to = "group", values_to = "value") %>%
    dplyr::arrange(taxon)

df_fig_trend$group = recode(df_fig_trend$group, 
                          `lfc1` = "Overweight - Obese",
                          `lfc2` = "Lean - Obese")
df_fig_trend$group = factor(df_fig_trend$group, 
                          levels = c("Overweight - Obese", 
                                     "Lean - Obese"))

lo = floor(min(df_fig_trend$value))
up = ceiling(max(df_fig_trend$value))
mid = (lo + up)/2
fig_trend = df_fig_trend %>%
    ggplot(aes(x = group, y = taxon, fill = value)) + 
    geom_tile(color = "black") +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         na.value = "white", midpoint = mid, limit = c(lo, up),
                         name = NULL) +
    geom_text(aes(group, taxon, label = value), color = "black", size = 4) +
    labs(x = NULL, y = NULL, title = "Log fold changes as compared to obese subjects") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.y = element_text(color = df_fig_trend %>%
                                         dplyr::distinct(taxon, color) %>%
                                         .$color))
fig_trend
```

## 4.10 Alternative input formats

The call in Section 4.2 also accepts a `TreeSummarizedExperiment` object or a
count matrix. Only the data arguments differ; all other arguments are unchanged.

For a `TreeSummarizedExperiment` object, pass `data = tse` and
`assay_name = "counts"`.

```{r, eval=FALSE}
tse = mia::convertFromPhyloseq(atlas1006)
tse = tse[, tse$time == 0]
tse$bmi = recode(tse$bmi_group,
                 obese = "obese",
                 severeobese = "obese",
                 morbidobese = "obese")
tse = tse[, tse$bmi %in% c("lean", "overweight", "obese")]
tse$bmi = factor(tse$bmi, levels = c("obese", "overweight", "lean"))
tse$region = recode(as.character(tse$nationality),
                    Scandinavia = "NE", UKIE = "NE", SouthEurope = "SE",
                    CentralEurope = "CE", EasternEurope = "EE",
                    .missing = "unknown")
tse = tse[, ! tse$region %in% c("EE", "unknown")]
```

For a count matrix, three objects are supplied: the abundance data at the
lowest available taxonomic level, the data aggregated to the taxonomic level of
the analysis, and the sample metadata. If no aggregation is performed, the
aggregated data are the same as the abundance data. Pass
`data = abundance_data`, `aggregate_data = aggregate_data`, and
`meta_data = meta_data`, and drop `tax_level`.

```{r, eval=FALSE}
abundance_data = microbiome::abundances(pseq)
aggregate_data = microbiome::abundances(microbiome::aggregate_taxa(pseq, "Family"))
meta_data = microbiome::meta(pseq)
```

# 5. Run ANCOM-BC2 on a real longitudinal dataset {.tabset}

## 5.1 Import example data

A two-week diet swap study between western (USA) and traditional (rural Africa)
diets [@lahti2014tipping]. The dataset is available via the
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq]
format.

```{r}
data(dietswap, package = "microbiome")
```

## 5.2 Run `ancombc2` function using the `phyloseq` object

In this tutorial, we consider the following fixed effects:

* Continuous covariates: "timepoint"

* Categorical covariates: "nationality"

* The group variable of interest: "group"

    + Three groups: "DI", "ED", "HE"

    + The reference group: "DI"

and the following random effects:

* A random intercept

* A random slope: "timepoint"

The ANCOM-BC2 global test, pairwise directional test, Dunnett's type of test,
and pattern analysis are the same as for the cross-sectional data.

```{r}
set.seed(123)
# It should be noted that we have set the number of bootstrap samples (B) equal 
# to 10 in the 'trend_control' function for computational expediency. 
# However, it is recommended that users utilize the default value of B, 
# which is 100, or larger values for optimal performance.
output = ancombc2(data = dietswap, tax_level = "Family",
                  fix_formula = "nationality + timepoint + group",
                  rand_formula = "(timepoint | subject)",
                  p_adj_method = "holm", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "group", struc_zero = TRUE, neg_lb = TRUE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
                  iter_control = list(tol = 1e-2, max_iter = 20, 
                                      verbose = TRUE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
                  trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
                                                              nrow = 2, 
                                                              byrow = TRUE)),
                                       node = list(2),
                                       B = 10))
```

Tests for interactions are again supported by specifying the interactions in
`fix_formula`, with the `*` operator.

```{r, eval=FALSE}
set.seed(123)
output = ancombc2(data = dietswap, tax_level = "Family",
                  fix_formula = "nationality + timepoint * group",
                  rand_formula = "(timepoint | subject)",
                  p_adj_method = "holm", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "group", struc_zero = TRUE, neg_lb = TRUE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
                  iter_control = list(tol = 1e-2, max_iter = 20, 
                                      verbose = TRUE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
                  trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
                                                              nrow = 2, 
                                                              byrow = TRUE)),
                                       node = list(2),
                                       B = 10))
```

## 5.3 ANCOM-BC2 primary analysis

```{r}
res_prim = output$res %>%
    mutate_if(is.numeric, function(x) round(x, 2))
res_prim[1:6, ] %>%
    datatable(caption = "ANCOM-BC2 Primary Analysis")
```

## 5.4 ANCOM-BC2 global test

```{r}
res_global = output$res_global %>%
    mutate_if(is.numeric, function(x) round(x, 2))
res_global[1:6, ] %>%
    datatable(caption = "ANCOM-BC2 Global Test")
```

## 5.5 ANCOM-BC2 multiple pairwise comparisons

```{r}
res_pair = output$res_pair %>%
    mutate_if(is.numeric, function(x) round(x, 2))
res_pair[1:6, ] %>%
    datatable(caption = "ANCOM-BC2 Multiple Pairwise Comparisons")
```

## 5.6 ANCOM-BC2 Dunnett's type of test

```{r}
res_dunn = output$res_dunn %>%
    mutate_if(is.numeric, function(x) round(x, 2))
res_dunn[1:6, ] %>%
    datatable(caption = "ANCOM-BC2 Dunnett's Type of Test")
```

## 5.7 ANCOM-BC2 pattern analysis

```{r}
res_trend = output$res_trend %>%
    mutate_if(is.numeric, function(x) round(x, 2))
res_trend[1:6, ] %>%
    datatable(caption = "ANCOM-BC2 Pattern Analysis")
```

## 5.8 Alternative input formats

The call in Section 5.2 also accepts a `TreeSummarizedExperiment` object or a
count matrix. Only the data arguments differ; all other arguments are unchanged.

For a `TreeSummarizedExperiment` object, pass `data = tse` and
`assay_name = "counts"`.

```{r, eval=FALSE}
tse = mia::makeTreeSummarizedExperimentFromPhyloseq(dietswap)
```

For a count matrix, pass `data = abundance_data`,
`aggregate_data = aggregate_data`, and `meta_data = meta_data`, and drop
`tax_level`.

```{r, eval=FALSE}
abundance_data = microbiome::abundances(dietswap)
aggregate_data = microbiome::abundances(microbiome::aggregate_taxa(dietswap, "Family"))
meta_data = microbiome::meta(dietswap)
```

# 6. Bias-corrected log abundances

The estimation of sampling fractions in ANCOM-BC2 is limited to an additive
constant, so that only differences between bias-corrected log abundances are
interpretable, and not the values themselves.

Within each taxon, the bias-corrected log abundances are centered across
samples. ANCOM-BC2 assumes that the taxon-specific biases differ between taxa
but stay constant within a taxon across samples, and the centering removes them.

```{r}
bias_correct_log_table = output$bias_correct_log_table
# By default, ANCOM-BC2 does not add pseudo-counts to zero counts, which can 
# result in NAs in the bias-corrected log abundances. Users have the option to 
# either leave the NAs as they are or replace them with zeros. 
# This replacement is equivalent to adding pseudo-counts of ones to the zero counts. 
bias_correct_log_table[is.na(bias_correct_log_table)] = 0
# Show the first 6 samples
round(bias_correct_log_table[, 1:6], 2) %>% 
  datatable(caption = "Bias-corrected log abundances")
```

# 7. Run ANCOM-BC2 on a simulated dataset {.tabset}

The remaining two sections evaluate the operating characteristics of ANCOM-BC2:
power and FDR on simulated data with known differentially abundant taxa, and
the distribution of the p-values on a dataset with a permuted group label.

## 7.1 Generate simulated data

1. The simulated data contain one continuous covariate, cont_cov, and one
categorical covariate, cat_cov, with three levels: "1", "2", and "3". The group
variable is cat_cov.

2. The true abundances are generated using the Poisson lognormal (PLN)
model based on the mechanism described in the LDM paper [@hu2020testing]. The
PLN model relates the abundance vector $Y_i$ with a Gaussian latent vector
$Z_i$ for taxon $i$ as follows:
$$ \text{latent layer: } Z_i \sim N(\mu_i, \Sigma_i) \\
\text{observation layer: } Y_i|Z_i \sim POI(N \exp(Z_i)) $$
where $N$ is the scaling factor. Because of the presence of a latent layer, the
PLN model displays a larger variance than the Poisson model (over-dispersion).
Also, the covariance (correlation) between abundances has the same sign as
the covariance (correlation) between the corresponding latent variables.
This property gives enormous flexibility in modeling the variance-covariance
structure of microbial abundances since it is easy to specify different
variance-covariance matrices in the multivariate Gaussian distribution.

3. Instead of specifying the variance-covariance matrix, we choose to estimate
the variance-covariance matrix from a real dataset: the Quantitative Microbiome
Project (QMP) data [@vandeputte2017quantitative]. This dataset contains
quantitative microbiome count data of 106 samples and 91 OTUs.

```{r}
data(QMP, package = "ANCOMBC")
set.seed(123)
n = 150
d = ncol(QMP)
diff_prop = 0.1
lfc_cont = 1
lfc_cat2_vs_1 = -2
lfc_cat3_vs_1 = 1

# Generate the true abundances
abn_data = sim_plnm(abn_table = QMP, taxa_are_rows = FALSE, prv_cut = 0.05, 
                    n = n, lib_mean = 1e8, disp = 0.5)
log_abn_data = log(abn_data + 1e-5)
rownames(log_abn_data) = paste0("T", seq_len(d))
colnames(log_abn_data) = paste0("S", seq_len(n))

# Generate the sample and feature meta data
# Sampling fractions are set to differ by batches
smd = data.frame(samp_frac = log(c(runif(n/3, min = 1e-4, max = 1e-3),
                                   runif(n/3, min = 1e-3, max = 1e-2),
                                   runif(n/3, min = 1e-2, max = 1e-1))),
                 cont_cov = rnorm(n),
                 cat_cov = as.factor(rep(seq_len(3), each = n/3)))
rownames(smd) = paste0("S", seq_len(n))
                      
fmd = data.frame(taxon = paste0("T", seq_len(d)),
                 seq_eff = log(runif(d, min = 0.1, max = 1)),
                 lfc_cont = sample(c(0, lfc_cont), 
                                   size = d,
                                   replace = TRUE,
                                   prob = c(1 - diff_prop, diff_prop)),
                 lfc_cat2_vs_1 = sample(c(0, lfc_cat2_vs_1), 
                                        size = d,
                                        replace = TRUE,
                                        prob = c(1 - diff_prop, diff_prop)),
                 lfc_cat3_vs_1 = sample(c(0, lfc_cat3_vs_1), 
                                        size = d,
                                        replace = TRUE,
                                        prob = c(1 - diff_prop, diff_prop))) %>%
    mutate(lfc_cat3_vs_2 = lfc_cat3_vs_1 - lfc_cat2_vs_1)

# Add effect sizes of covariates to the true abundances
smd_dmy = model.matrix(~ 0 + cont_cov + cat_cov, data = smd)
log_abn_data = log_abn_data + outer(fmd$lfc_cont, smd_dmy[, "cont_cov"] )
log_abn_data = log_abn_data + outer(fmd$lfc_cat2_vs_1, smd_dmy[, "cat_cov2"])
log_abn_data = log_abn_data + outer(fmd$lfc_cat3_vs_1, smd_dmy[, "cat_cov3"])

# Add sample- and taxon-specific biases
log_otu_data = t(t(log_abn_data) + smd$samp_frac)
log_otu_data = log_otu_data + fmd$seq_eff
otu_data = round(exp(log_otu_data))
```

## 7.2 Run `ancombc2` function

```{r}
set.seed(123)
output = ancombc2(data = otu_data, meta_data = smd, 
                  fix_formula = "cont_cov + cat_cov", rand_formula = NULL,
                  p_adj_method = "holm", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "cat_cov", struc_zero = FALSE, neg_lb = FALSE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = FALSE, pairwise = TRUE, 
                  dunnet = FALSE, trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, 
                                      verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = NULL, 
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

res_prim = output$res
res_pair = output$res_pair
```

## 7.3 Power and FDR

A taxon is counted as a true positive for a pairwise comparison when the sign of
its estimated log fold change agrees with the sign of the true log fold change
and the taxon is declared significant. The accounting is carried out twice,
once with the `diff_` indicators and once with the `diff_robust_` indicators.

```{r}
pair_col = c("cat_cov2", "cat_cov3", "cat_cov3_cat_cov2")

# True direction of the log fold change of each pairwise comparison
lfc_true = fmd %>%
    dplyr::transmute(taxon,
                     true_cat_cov2 = sign(lfc_cat2_vs_1),
                     true_cat_cov3 = sign(lfc_cat3_vs_1),
                     true_cat_cov3_cat_cov2 = sign(lfc_cat3_vs_2))

# Power and FDR of the pairwise directional test, with significance declared by
# the indicator columns carrying the given prefix
power_fdr = function(prefix) {
    df = res_pair %>%
        dplyr::left_join(lfc_true, by = "taxon")
    tp = 0
    fp = 0
    fn = 0
    for (x in pair_col) {
        est = sign(df[[paste0("lfc_", x)]] * df[[paste0(prefix, x)]])
        est[is.na(est)] = 0
        true = df[[paste0("true_", x)]]
        true[is.na(true)] = 0
        tp = tp + sum(true == 1 & est == 1) + sum(true == -1 & est == -1)
        fp = fp + sum(true == 0 & est != 0) +
            sum(true == 1 & est == -1) + sum(true == -1 & est == 1)
        fn = fn + sum(true != 0 & est == 0)
    }
    c(power = tp/(tp + fn), fdr = fp/(tp + fp))
}

acc_diff = power_fdr("diff_")
acc_robust = power_fdr("diff_robust_")

tab_summ = data.frame(Comparison = c("Without sensitivity score filter",
                                     "With sensitivity score filter"),
                      Power = round(unname(c(acc_diff["power"],
                                             acc_robust["power"])), 2),
                      FDR = round(unname(c(acc_diff["fdr"],
                                           acc_robust["fdr"])), 2))
tab_summ %>%
    datatable(caption = "Power/FDR Comparison")
```

## 7.4 Sensitivity analysis with `conservative = FALSE`

The `conservative` argument sets the stage at which the pseudo-counts are added.
Let $\alpha$ denote the significance level.

Under the default, `conservative = TRUE`, the pseudo-counts
$c \in \{0, 0.1, 0.5, 1\}$ are added to all counts before the sampling
fractions are estimated, and the full algorithm is re-run for each $c$. Let
$q_c$ denote the adjusted p-value of a taxon in run $c$. The sensitivity score
of the taxon is
$$ s = \frac{1}{4} \sum_{c \in \{0,\, 0.1,\, 0.5,\, 1\}} \mathbf{1}\{q_c > \alpha\}, $$
and the taxon passes the sensitivity analysis when
$$ \text{passed\_ss} = \mathbf{1}\{s = 0 \text{ or } s = 1\} = 1. $$

Under `conservative = FALSE`, the sampling fractions are estimated once on the
complete data, the pseudo-counts $c \in \{0.01, 0.02, \ldots, 0.50\}$ are added
to the zero counts of the bias-corrected data, and the fixed effects are
re-estimated by ordinary least squares for each $c$. Let $p_c$ denote the
unadjusted p-value of refit $c$, and $p_0$ the unadjusted p-value of the main
run. The sensitivity score of the taxon is
$$ s = \frac{1}{50} \sum_{c \in \{0.01,\, 0.02,\, \ldots,\, 0.50\}} \mathbf{1}\{p_c > \alpha\}, $$
and the taxon passes the sensitivity analysis when
$$ \text{passed\_ss} = \mathbf{1}\{(s = 0 \text{ and } p_0 \le \alpha) \text{ or } (s = 1 \text{ and } p_0 > \alpha)\} = 1. $$

Under both settings, a taxon is declared robust when it is differentially
abundant and passes the sensitivity analysis,
$$ \text{diff\_robust} = \text{diff\_abn} \wedge \text{passed\_ss}. $$

Under `conservative = TRUE` a taxon that is nonsignificant in every run also
satisfies $s = 1$ and therefore passes, so `passed_ss` alone is not an indicator
of significance. The default is recommended unless power is the primary concern:
`conservative = FALSE` has higher power and a higher false positive rate.

```{r}
set.seed(123)
output_sens = ancombc2(data = otu_data, meta_data = smd,
                       fix_formula = "cont_cov + cat_cov", rand_formula = NULL,
                       p_adj_method = "holm", pseudo_sens = TRUE,
                       conservative = FALSE,
                       prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                       group = "cat_cov", struc_zero = FALSE, neg_lb = FALSE,
                       alpha = 0.05, n_cl = 2, verbose = FALSE,
                       global = FALSE, pairwise = TRUE,
                       dunnet = FALSE, trend = FALSE,
                       iter_control = list(tol = 1e-5, max_iter = 20,
                                           verbose = FALSE),
                       em_control = list(tol = 1e-5, max_iter = 100),
                       lme_control = NULL,
                       mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
                       trend_control = NULL)

res_pair_sens = output_sens$res_pair

pair_col = c("cat_cov2", "cat_cov3", "cat_cov3_cat_cov2")
tab_sens = data.frame(
    Comparison = pair_col,
    Significant = colSums(res_pair[, paste0("diff_", pair_col)]),
    `Robust (conservative = TRUE)` =
        colSums(res_pair[, paste0("diff_robust_", pair_col)]),
    `Robust (conservative = FALSE)` =
        colSums(res_pair_sens[, paste0("diff_robust_", pair_col)]),
    `Passed (conservative = TRUE)` =
        colSums(res_pair[, paste0("passed_ss_", pair_col)]),
    `Passed (conservative = FALSE)` =
        colSums(res_pair_sens[, paste0("passed_ss_", pair_col)]),
    row.names = NULL, check.names = FALSE)
tab_sens %>%
    datatable(caption = "Sensitivity Analysis Comparison")
```

The two settings declare the same taxa robust in the pairwise
directional test on this dataset:
`r paste(tab_sens[["Robust (conservative = TRUE)"]], collapse = ", ")` taxa for
the three comparisons. They differ in the sensitivity analysis itself. Of the
`r nrow(res_pair)` taxa,
`r paste(tab_sens[["Passed (conservative = FALSE)"]], collapse = ", ")` are
declared insensitive to the pseudo-count addition under `conservative = FALSE`,
against
`r paste(tab_sens[["Passed (conservative = TRUE)"]], collapse = ", ")` under the
default. The taxa whose sensitivity flags differ between the two settings are
not significant in the pairwise directional test.

# 8. Benchmark the performance of ANCOM-BC2 on a permuted null dataset

## 8.1 Permute the `bmi` label

The `bmi` label of the HITChip Atlas dataset imported in Section 4.1 is
permuted across samples. After the permutation the data are null for `bmi`: a
method that controls false positives produces approximately uniform p-values.

```{r}
set.seed(123)

pseq_perm = pseq
meta_data_perm = microbiome::meta(pseq_perm)
meta_data_perm$bmi = sample(meta_data_perm$bmi)

phyloseq::sample_data(pseq_perm) = meta_data_perm
```

## 8.2 Run `ancombc2` function 

```{r}
set.seed(123)
output = ancombc2(data = pseq_perm, tax_level = "Genus",
                  fix_formula = "bmi", rand_formula = NULL,
                  p_adj_method = "holm", pseudo_sens = TRUE,
                  prv_cut = 0, lib_cut = 1000, s0_perc = 0.05,
                  group = "bmi", struc_zero = TRUE, neg_lb = TRUE)
```

## 8.3 Visualization

The p-values of ANCOM-BC2 on the permuted data are close to uniform, with and
without the sensitivity analysis. Discarding the significant p-values of the
taxa that fail the sensitivity analysis removes part of the excess near zero and
therefore lowers the number of false positives. We recommend including the
sensitivity analysis in the assessment of taxon significance, and taking it into
account when ANCOM-BC2 is used as a comparator in benchmarks of false positive
control.

```{r}
res_perm = output$res %>%
    dplyr::transmute(taxon = taxon,
                     p = p_bmioverweight,
                     ss_pass = passed_ss_bmioverweight) %>%
  rowwise() %>%
  mutate(
      p_update = case_when(
          # For taxa with significant p-values that failed the sensitivity analysis,
          # we assign a random value uniformly drawn from the range [0.05, 1].
          p < 0.05 & ss_pass == FALSE ~ runif(1, min = 0.05, max = 1),
    TRUE ~ p 
  )) %>%
  ungroup()

df_p = res_perm %>%
  dplyr::select(p, p_update) %>%
  pivot_longer(cols = p:p_update, names_to = "type", values_to = "value") 
df_p$type = recode(df_p$type, 
                   `p` = "ANCOM-BC2 (Without Sensitivity Analysis)", 
                   `p_update` = "ANCOM-BC2 (With Sensitivity Analysis)")

num_bins = 30
fig_p = df_p %>%
  ggplot(aes(x = value, fill = type)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = num_bins) +
  geom_hline(yintercept = phyloseq::ntaxa(pseq_perm)/num_bins, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Distribution of P-values",
       x = "P-value",
       y = "Frequency",
       fill = NULL,
       color = NULL) +
    theme_bw() + 
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.minor.y = element_blank(),
          legend.position = "bottom")
fig_p
```

# Session information

```{r sessionInfo, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```

# References
