---
title: "ANCOM-BC"
author: 
  - Huang Lin$^1$
  - $^1$NICHD, 6710B Rockledge Dr, Bethesda, MD 20892
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: 
    toc: true
    theme: united
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{ANCOMBC}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA, 
                      fig.width = 6.25, fig.height = 5)
library(ANCOMBC)
library(tidyverse)
library(microbiome)
library(magrittr)
library(qwraps2)
library(corrplot)
library(limma)
library(ggforce)
library(DT)
options(DT.options = list(
  initComplete = JS("function(settings, json) {",
  "$(this.api().table().header()).css({'background-color': 
  '#000', 'color': '#fff'});","}")))
```

# 1. Introduction

Analysis of Compositions of Microbiomes with Bias Correction (ANCOM-BC) [@lin2020analysis] 
is a methodology of differential abundance (DA) analysis for microbial absolute
abundances. ANCOM-BC estimates the unknown sampling fractions, corrects 
the bias induced by their differences through a log linear regression model 
including the estimated sampling fraction as an offset terms, and identifies 
taxa that are differentially abundant according to the variable of interest. 
For more details, please refer to the 
[ANCOM-BC](https://doi.org/10.1038/s41467-020-17041-7) paper.

Analysis of Composition of Microbiomes (ANCOM) [@mandal2015analysis] is also a 
DA analysis for microbial absolute abundances. It accounts for the 
compositionality of microbiome data by performing the additive log ratio (ALR) 
transformation. ANCOM employs a heuristic strategy to declare taxa that are 
significantly differentially abundant. For a given taxon, the output W statistic 
represents the number ALR transformed models where the taxon is differentially 
abundant with regard to the variable of interest. Larger the value of W, the 
more likely the taxon is differentially abundant. For more details, 
please refer to the 
[ANCOM](https://www.tandfonline.com/doi/full/10.3402/mehd.v26.27663) paper.

Sparse Estimation of Correlations among Microbiomes (SECOM) is a methodology
which aims to detect both linear and nonlinear relationships between a pair of 
taxa within an ecosystem (e.g. gut) or across ecosystems (e.g. gut and tongue).
SECOM corrects both sample-specific and taxon-specific biases, obtains a
consistent estimator for the correlation matrix of microbial absolute 
abundances, while maintaining the underlying true sparsity.

# 2. Installation

Download package. 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ANCOMBC")
```

Load the package. 

```{r load, eval=FALSE}
library(ANCOMBC)
```

# 3. Example Data {.tabset}

## Cross-sectional data

The HITChip Atlas data set [@lahti2014tipping] is available via the 
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq] 
format. 

```{r data1}
data(atlas1006) 
# Subset to baseline
pseq = subset_samples(atlas1006, time == 0)

# Re-code the bmi group
sample_data(pseq)$bmi_group = recode(sample_data(pseq)$bmi_group,
                                     lean = "lean",
                                     overweight = "overweight",
                                     obese = "obese",
                                     severeobese = "obese",
                                     morbidobese = "obese")
# Subset to lean and obese subjects
pseq = subset_samples(pseq, bmi_group %in% c("lean", "obese"))

# Create the region variable
sample_data(pseq)$region = recode(sample_data(pseq)$nationality,
                                  Scandinavia = "NE",
                                  UKIE = "NE",
                                  SouthEurope = "SE",
                                  CentralEurope = "CE",
                                  EasternEurope = "EE")

# Discard "EE" as it contains only 1 subject
pseq = subset_samples(pseq, region != "EE")
# Genus level data
genus_data = aggregate_taxa(pseq, "Genus")
# Family level data
family_data = aggregate_taxa(pseq, "Family")
# Phylum level data
phylum_data = aggregate_taxa(pseq, "Phylum")

print(genus_data)
print(family_data)
print(phylum_data)
```

## Longitudinal data

A two-week diet swap study between western (USA) and traditional (rural Africa) 
diets [@lahti2014tipping]. The data set is available via the 
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq] 
format.

```{r data2}
data(dietswap)

# Aggregate to family level
family_data2 = aggregate_taxa(dietswap, "Family")

print(family_data2)
```

# 4. ANCOMBC Implementation

## 4.1 Run ancombc function

```{r ancombc}
out = ancombc(phyloseq = family_data, formula = "age + region + bmi_group", 
              p_adj_method = "holm", prv_cut = 0.10, lib_cut = 1000, 
              group = "region", struc_zero = TRUE, neg_lb = TRUE, tol = 1e-5, 
              max_iter = 100, conserve = TRUE, alpha = 0.05, global = TRUE)

res = out$res
res_global = out$res_global
```

## 4.2 ANCOMBC primary result {.tabset}

Result from the ANCOM-BC log-linear model to determine taxa that are 
differentially abundant according to the covariate of interest. It contains: 
1) log fold changes; 2) standard errors; 3) test statistics; 4) p-values; 
5) adjusted p-values; 6) indicators whether the taxon is differentially 
abundant (TRUE) or not (FALSE).

### LFC

```{r}
tab_lfc = res$lfc
col_name = c("Age", "NE - CE", "SE - CE", "US - CE", "Obese - Lean")
colnames(tab_lfc) = col_name
tab_lfc %>% 
  datatable(caption = "Log Fold Changes from the Primary Result") %>%
  formatRound(col_name, digits = 2)
```

### SE

```{r}
tab_se = res$se
colnames(tab_se) = col_name
tab_se %>% 
  datatable(caption = "SEs from the Primary Result") %>%
  formatRound(col_name, digits = 2)
```

### Test statistic

```{r}
tab_w = res$W
colnames(tab_w) = col_name
tab_w %>% 
  datatable(caption = "Test Statistics from the Primary Result") %>%
  formatRound(col_name, digits = 2)
```

### P-values

```{r}
tab_p = res$p_val
colnames(tab_p) = col_name
tab_p %>% 
  datatable(caption = "P-values from the Primary Result") %>%
  formatRound(col_name, digits = 2)
```

### Adjusted p-values

```{r}
tab_q = res$q
colnames(tab_q) = col_name
tab_q %>% 
  datatable(caption = "Adjusted p-values from the Primary Result") %>%
  formatRound(col_name, digits = 2)
```

### Differentially abundant taxa

```{r}
tab_diff = res$diff_abn
colnames(tab_diff) = col_name
tab_diff %>% 
  datatable(caption = "Differentially Abundant Taxa from the Primary Result")
```

### Visualization for age

```{r}
df_lfc = data.frame(res$lfc * res$diff_abn, check.names = FALSE) %>% 
  rownames_to_column("taxon_id")
df_se = data.frame(res$se * res$diff_abn, check.names = FALSE) %>% 
  rownames_to_column("taxon_id")
colnames(df_se)[-1] = paste0(colnames(df_se)[-1], "SE")

df_fig_age = df_lfc %>% 
  dplyr::left_join(df_se, by = "taxon_id") %>%
  dplyr::transmute(taxon_id, age, ageSE) %>%
  dplyr::filter(age != 0) %>% 
  dplyr::arrange(desc(age)) %>%
  dplyr::mutate(direct = ifelse(age > 0, "Positive LFC", "Negative LFC"))
df_fig_age$taxon_id = factor(df_fig_age$taxon_id, levels = df_fig_age$taxon_id)
df_fig_age$direct = factor(df_fig_age$direct, 
                        levels = c("Positive LFC", "Negative LFC"))
  
p_age = ggplot(data = df_fig_age, 
           aes(x = taxon_id, y = age, fill = direct, color = direct)) + 
  geom_bar(stat = "identity", width = 0.7, 
           position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = age - ageSE, ymax = age + ageSE), width = 0.2,
                position = position_dodge(0.05), color = "black") + 
  labs(x = NULL, y = "Log fold change", 
       title = "Waterfall Plot of Age") + 
  scale_fill_discrete(name = NULL) +
  scale_color_discrete(name = NULL) +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1))
p_age
```

### Visualization for BMI

```{r}
df_fig_bmi = df_lfc %>% 
  dplyr::left_join(df_se, by = "taxon_id") %>%
  dplyr::transmute(taxon_id, bmi = bmi_groupobese, bmiSE = bmi_groupobeseSE) %>%
  dplyr::filter(bmi != 0) %>% 
  dplyr::arrange(desc(bmi)) %>%
  dplyr::mutate(direct = ifelse(bmi > 0, "Positive LFC", "Negative LFC"))
df_fig_bmi$taxon_id = factor(df_fig_bmi$taxon_id, levels = df_fig_bmi$taxon_id)
df_fig_bmi$direct = factor(df_fig_bmi$direct, 
                        levels = c("Positive LFC", "Negative LFC"))
  
p_bmi = ggplot(data = df_fig_bmi, 
           aes(x = taxon_id, y = bmi, fill = direct, color = direct)) + 
  geom_bar(stat = "identity", width = 0.7, 
           position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = bmi - bmiSE, ymax = bmi + bmiSE), width = 0.2,
                position = position_dodge(0.05), color = "black") + 
  labs(x = NULL, y = "Log fold change", 
       title = "Waterfall Plot of BMI (Obese - Lean)") + 
  scale_fill_discrete(name = NULL) +
  scale_color_discrete(name = NULL) +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1))
p_bmi
```

## 4.3 ANCOMBC global test result {.tabset}

Result from the ANCOM-BC global test to determine taxa that are 
differentially abundant between at least two groups across three or more 
different groups. In this example, we want to identify taxa that are 
differentially abundant between at least two regions across CE, NE, SE, and US.
The result contains: 1) test statistics; 2) p-values; 3) adjusted p-values; 
4) indicators whether the taxon is differentially abundant (TRUE) or not 
(FALSE).

### Test statistics

```{r}
tab_w = res_global[, "W", drop = FALSE]
colnames(tab_w) = "Region"
tab_w %>% datatable(caption = "Test Statistics 
                    from the Global Test Result") %>%
      formatRound(c("Region"), digits = 2)
```

### P-values

```{r}
tab_p = res_global[, "p_val", drop = FALSE]
colnames(tab_p) = "Region"
tab_p %>% datatable(caption = "P-values 
                    from the Global Test Result") %>%
      formatRound(c("Region"), digits = 2)
```

### Adjusted p-values

```{r}
tab_q = res_global[, "q_val", drop = FALSE]
colnames(tab_q) = "Region"
tab_q %>% datatable(caption = "Adjusted p-values 
                    from the Global Test Result") %>%
      formatRound(c("Region"), digits = 2)
```

### Differentially abundant taxa

```{r}
tab_diff = res_global[, "diff_abn", drop = FALSE]
colnames(tab_diff) = "Region"
tab_diff %>% datatable(caption = "Differentially Abundant Taxa 
                       from the Global Test Result")
```

### Bias-corrected abundances

Step 1: estimate sample-specific sampling fractions (in log scale). Note that
for each sample, if it contains missing values for any variable specified in the 
formula, the corresponding sampling fraction estimate for this sample will 
be NA since the sampling fraction is not estimable with the presence of 
missing values.

Step 2: correct the log observed abundances by subtracting the estimated 
sampling fraction from log observed abundances of each sample.

```{r}
samp_frac = out$samp_frac
# Replace NA with 0
samp_frac[is.na(samp_frac)] = 0 

# Add pesudo-count (1) to avoid taking the log of 0
log_obs_abn = log(abundances(family_data) + 1) 
# Adjust the log observed abundances
log_obs_abn_adj = t(t(log_obs_abn) - samp_frac)
# Show the first 6 samples
round(log_obs_abn_adj[, 1:6], 2) %>% 
  datatable(caption = "Bias-corrected log observed abundances")
```

### Visualization

```{r}
sig_taxa = res_global %>%
  tibble::rownames_to_column("taxon") %>%
  dplyr::filter(diff_abn == TRUE) %>%
  .$taxon

df_sig = as.data.frame(t(log_obs_abn_adj[sig_taxa, ])) %>%
  tibble::rownames_to_column("sample") %>%
  dplyr::left_join(meta(family_data) %>%
                     select(sample, region),
                   by = "sample") %>%
  dplyr::filter(!is.na(region)) %>%
  tidyr::pivot_longer(cols = -one_of("sample", "region"), 
                      names_to = "taxon", values_to = "value")

df_heat = df_sig %>%
  dplyr::group_by(region, taxon) %>%
  dplyr::summarise_if(is.numeric, mean, na.rm = TRUE) %>%
  dplyr::mutate(value = round(value, 2)) %>%
  dplyr::arrange(region)
df_heat$taxon = factor(df_heat$taxon, levels = sig_taxa)

lo = floor(min(df_heat$value))
up = ceiling(max(df_heat$value))
mid = (lo + up)/2
p_heat = df_heat %>%
  ggplot(aes(x = region, y = taxon, fill = value)) + 
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       na.value = "white", midpoint = mid, limit = c(lo, up),
                       name = NULL) +
  geom_text(aes(region, taxon, label = value), color = "black", size = 4) +
  labs(x = NULL, y = NULL, title = "Heat Map of Region") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
p_heat
```

# 5. ANCOM Implementation

## 5.1 Cross-sectional data {.tabset}

### Run ancom function

```{r}
set.seed(123)
out1 = ancom(phyloseq = family_data,  p_adj_method = "holm",
             prv_cut = 0.10, lib_cut = 1000, main_var = "bmi_group", 
             adj_formula = "age + region", rand_formula = NULL,
             lme_control = NULL, struc_zero = TRUE, neg_lb = TRUE,
             alpha = 0.05, n_cl = 2)

res1 = out1$res

# Similarly, if the main variable of interest is continuous, such as age, the
# ancom model can be specified as
# out1 = ancom(phyloseq = family_data,  p_adj_method = "holm",
#              prv_cut = 0.10, lib_cut = 0, main_var = "age",
#              adj_formula = "region + bmi_group", rand_formula = NULL,
#              lme_control = NULL, struc_zero = FALSE, neg_lb = FALSE,
#              alpha = 0.05, n_cl = 2)
```

### Visualization for W statistics

```{r}
q_val = out1$q_data
beta_val = out1$beta_data
# Only consider the effect sizes with the corresponding q-value less than alpha
beta_val = beta_val * (q_val < 0.05) 
# Choose the maximum of beta's as the effect size
beta_pos = apply(abs(beta_val), 2, which.max) 
beta_max = vapply(seq_along(beta_pos), function(i) beta_val[beta_pos[i], i],
                  FUN.VALUE = double(1))
# Number of taxa except structural zeros
n_taxa = ifelse(is.null(out1$zero_ind), 
                nrow(feature_table), 
                sum(apply(out1$zero_ind, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = 0.7 * (n_taxa - 1)

df_fig_w = res1 %>%
  dplyr::mutate(beta = beta_max,
                direct = case_when(
                  detected_0.7 == TRUE & beta > 0 ~ "Positive",
                  detected_0.7 == TRUE & beta <= 0 ~ "Negative",
                  TRUE ~ "Not Significant"
                  )) %>%
  dplyr::arrange(W)
df_fig_w$taxon_id = factor(df_fig_w$taxon_id, levels = df_fig_w$taxon_id)
df_fig_w$W = replace(df_fig_w$W, is.infinite(df_fig_w$W), n_taxa - 1)
df_fig_w$direct = factor(df_fig_w$direct, 
                     levels = c("Negative", "Positive", "Not Significant"))

p_w = df_fig_w %>%
  ggplot(aes(x = taxon_id, y = W, color = direct)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(x = "Taxon", y = "W") +
  scale_color_discrete(name = NULL) + 
  geom_hline(yintercept = cut_off, linetype = "dotted", 
             color = "blue", size = 1.5) +
  geom_text(aes(x = 2, y = cut_off + 0.5, label = "W[0.7]"), 
            size = 5, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank())
p_w
```

### Compare ANCOM-BC and ANCOM results

```{r}
df_compare = res$diff_abn %>%
  tibble::rownames_to_column("taxon_id") %>%
  dplyr::left_join(res1, by = "taxon_id") %>%
  dplyr::transmute(ancombc = bmi_groupobese,
                   ancom = detected_0.7)
vdc = vennCounts(df_compare)
class(vdc) = "matrix"
df_vdc = as.data.frame(vdc)[-1, ] %>%
  dplyr::mutate(x = c(1, -1, 0),
                y = c(0, 0, 0)) %>%
  dplyr::transmute(x, y, label = Counts)

df_fig_venn = data.frame(x = c(-0.866, 0.866),
                     y = c(0, 0),
                     labels = c("ANCOM-BC", "ANCOM"))
df_fig_venn$labels = factor(df_fig_venn$labels, 
                            levels = c("ANCOM-BC", "ANCOM"))
p_venn = df_fig_venn %>%
  ggplot(aes(x0 = x, y0 = y, r = 1.5, fill = labels)) +
  geom_circle(alpha = .3, size = 1, color = "grey") +
  geom_text(data = df_vdc, aes(x, y, label = label),
            inherit.aes = FALSE, size = 8)+
  labs(fill = NULL) +
  coord_fixed() +
  theme_void()
p_venn
```

## 5.2 Longitudinal data {.tabset}

### Run ancom function

```{r}
set.seed(123)
out2 = ancom(phyloseq = family_data2,  p_adj_method = "holm",
             prv_cut = 0.10, lib_cut = 1000, main_var = "group", 
             adj_formula = "nationality + timepoint + bmi_group", 
             rand_formula = "~ timepoint | subject",
             lme_control = list(maxIter = 100, msMaxIter = 100, opt = "optim"), 
             struc_zero = TRUE, neg_lb = TRUE, alpha = 0.05, n_cl = 2)

res2 = out2$res
```

### Visualization for W statistics

```{r}
q_val = out2$q_data
beta_val = out2$beta_data
# Only consider the effect sizes with the corresponding q-value less than alpha
beta_val = beta_val * (q_val < 0.05) 
# Choose the maximum of beta's as the effect size
beta_pos = apply(abs(beta_val), 2, which.max) 
beta_max = vapply(seq_along(beta_pos), function(i) beta_val[beta_pos[i], i],
                  FUN.VALUE = double(1))
# Number of taxa except structural zeros
n_taxa = ifelse(is.null(out2$zero_ind), 
                nrow(feature_table), 
                sum(apply(out2$zero_ind, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = 0.7 * (n_taxa - 1)

df_fig_w = res2 %>%
  dplyr::mutate(beta = beta_max,
                direct = case_when(
                  detected_0.7 == TRUE & beta > 0 ~ "Positive",
                  detected_0.7 == TRUE & beta <= 0 ~ "Negative",
                  TRUE ~ "Not Significant"
                  )) %>%
  dplyr::arrange(W)
df_fig_w$taxon_id = factor(df_fig_w$taxon_id, levels = df_fig_w$taxon_id)
df_fig_w$W = replace(df_fig_w$W, is.infinite(df_fig_w$W), n_taxa - 1)
df_fig_w$direct = factor(df_fig_w$direct, 
                     levels = c("Negative", "Positive", "Not Significant"))

p_w = df_fig_w %>%
  ggplot(aes(x = taxon_id, y = W, color = direct)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(x = "Taxon", y = "W") +
  scale_color_discrete(name = NULL) + 
  geom_hline(yintercept = cut_off, linetype = "dotted", 
             color = "blue", size = 1.5) +
  geom_text(aes(x = 2, y = cut_off + 0.5, label = "W[0.7]"), 
            size = 5, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank())
p_w
```

# 6. SECOM Implementation

## 6.1 Run secom functions

```{r secom}
set.seed(123)
# Linear relationships
res_linear = secom_linear(pseqs = list(c(genus_data, phylum_data)), pseudo = 0, 
                          prv_cut = 0.5, lib_cut = 1000, corr_cut = 0.5, 
                          wins_quant = c(0.05, 0.95), method = "pearson", 
                          soft = FALSE, thresh_len = 20, n_cv = 10, 
                          thresh_hard = 0.3, max_p = 0.005, n_cl = 2)

# Nonlinear relationships
res_dist = secom_dist(pseqs = list(c(genus_data, phylum_data)), pseudo = 0, 
                      prv_cut = 0.5, lib_cut = 1000, corr_cut = 0.5, 
                      wins_quant = c(0.05, 0.95), R = 1000, 
                      thresh_hard = 0.3, max_p = 0.005, n_cl = 2)
```

## 6.2 Visualizations {.tabset}

### Pearson correlation with thresholding

```{r}
corrplot(res_linear$corr_th, method = "color", addgrid.col = "grey",
         order = "alphabet", type = "lower", col.lim = c(-1, 1), 
         diag = FALSE, addCoef.col = "black", number.cex = 0.8,
         col = COL2("BrBG"), title = "Pearson (Thresholding)", 
         mar = c(1, 0, 1, 0))
```

### Pearson correlation with p-value filtering

```{r}
corrplot(res_linear$corr_fl, method = "color", addgrid.col = "grey",
         order = "alphabet", type = "lower", col.lim = c(-1, 1), 
         diag = FALSE, addCoef.col = "black", number.cex = 0.8,
         col = COL2("BrBG"), title = "Pearson (Filtering)", 
         mar = c(1, 0, 1, 0))
```

### Distance correlation with p-value filtering

```{r}
corrplot(res_dist$dcorr_fl, method = "color", addgrid.col = "grey",
         order = "alphabet", type = "lower", col.lim = c(-1, 1), 
         diag = FALSE, addCoef.col = "black", number.cex = 0.8,
         col = COL2("BrBG"), title = "Distance (Filtering)", 
         mar = c(1, 0, 1, 0))
```

# Session information

```{r sessionInfo, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```

# References









