---
title: "ANCOM Tutorial"
author: 
  - Huang Lin$^1$
  - $^1$University of Maryland, College Park, MD 20742
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{ANCOM Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, message = FALSE, warning = FALSE, comment = NA}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA,
                      fig.width = 6.25, fig.height = 5)
library(ANCOMBC)
library(tidyverse)
```

# 1. Introduction

Analysis of Composition of Microbiomes (ANCOM) [@mandal2015analysis] is a 
differential abundance (DA) analysis for microbial absolute abundances. 
It accounts for the compositionality of microbiome data by performing
the additive log ratio (ALR) transformation. ANCOM employs a heuristic strategy
to declare taxa that are significantly differentially abundant.
For a given taxon, the W statistic is the number of ALR transformed models in
which the taxon is differentially abundant with respect to the variable
of interest. The larger the value of W, the stronger the evidence that the
taxon is differentially abundant. For more details, see the
[ANCOM](https://www.tandfonline.com/doi/full/10.3402/mehd.v26.27663) paper.

# 2. Installation

Download package. 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ANCOMBC")
```

Load the package. 

```{r load, eval=FALSE}
library(ANCOMBC)
```

# 3. Run ANCOM on a real cross-sectional dataset {.tabset}

## 3.1 Import example data

The HITChip Atlas dataset contains genus-level microbiota profiling with 
HITChip for 1006 western adults with no reported health complications, 
reported in [@lahti2014tipping]. The dataset is available via the 
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq] 
format. In this tutorial, we consider the following covariates:

* Continuous covariates: "age"

* Categorical covariates: "region", "bmi"

* The group variable of interest: "bmi"

    + Three groups: "lean", "overweight", "obese"
    
    + The reference group: "obese"

```{r, child = "children/child_atlas1006.Rmd"}
```

## 3.2 Run `ancom` function using `phyloseq` data

```{r}
set.seed(123)
out = ancom(data = pseq, tax_level = "Family", meta_data = NULL,
            p_adj_method = "holm", prv_cut = 0.10,
            lib_cut = 1000, main_var = "bmi", adj_formula = "age + region", 
            rand_formula = NULL, lme_control = NULL, struc_zero = TRUE,
            neg_lb = TRUE, alpha = 0.05, n_cl = 2, verbose = TRUE)

res = out$res

# Similarly, if the main variable of interest is continuous, such as age, the
# ancom model can be specified as
# out = ancom(data = pseq, tax_level = "Family", meta_data = NULL,
#             p_adj_method = "holm", prv_cut = 0.10,
#             lib_cut = 1000, main_var = "age", adj_formula = "bmi + region",
#             rand_formula = NULL, lme_control = NULL, struc_zero = FALSE,
#             neg_lb = FALSE, alpha = 0.05, n_cl = 2, verbose = TRUE)
```

## 3.3 Scatter plot for W statistics

```{r}
q_val = out$q_data
beta_val = out$beta_data
# Only consider the effect sizes with the corresponding q-value less than alpha
beta_val = beta_val * (q_val < 0.05) 
# Choose the maximum of beta's as the effect size
beta_pos = apply(abs(beta_val), 2, which.max) 
beta_max = vapply(seq_along(beta_pos), function(i) 
    beta_val[beta_pos[i], i], FUN.VALUE = double(1))
# Number of taxa included in the ALR comparisons, i.e., the taxa left after
# filtering and after the removal of taxa with structural zeros
n_taxa = nrow(out$q_data)
# Cutoff values for declaring differentially abundant taxa
cut_off = 0.7 * (n_taxa - 1)

df_fig_w = res %>%
  dplyr::mutate(beta = beta_max,
                direct = case_when(
                  detected_0.7 == TRUE & beta > 0 ~ "Positive",
                  detected_0.7 == TRUE & beta <= 0 ~ "Negative",
                  TRUE ~ "Not Significant"
                  )) %>%
  dplyr::arrange(W)
df_fig_w$taxon = factor(df_fig_w$taxon, levels = df_fig_w$taxon)
df_fig_w$W = replace(df_fig_w$W, is.infinite(df_fig_w$W), n_taxa - 1)
df_fig_w$direct = factor(df_fig_w$direct, 
                         levels = c("Negative", "Positive", "Not Significant"))

p_w = df_fig_w %>%
  ggplot(aes(x = taxon, y = W, color = direct)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(x = "Taxon", y = "W") +
  scale_color_discrete(name = NULL) + 
  geom_hline(yintercept = cut_off, linetype = "dotted", 
             color = "blue", size = 1.5) +
  geom_text(aes(x = 2, y = cut_off + 0.5, label = "W[0.7]"), 
            size = 5, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank())
p_w
```

## 3.4 Alternative input formats

The call in Section 3.2 also accepts a `TreeSummarizedExperiment` object or a
count matrix. Only the data arguments differ; all other arguments are unchanged.

For a `TreeSummarizedExperiment` object, pass `data = tse` and
`assay_name = "counts"`.

```{r, eval=FALSE}
tse = mia::makeTreeSummarizedExperimentFromPhyloseq(pseq)
```

For a count matrix, aggregation to the analysis level is performed beforehand.
Pass `data = abundance_data`, `aggregate_data = aggregate_data`, and
`meta_data = meta_data`, and drop `tax_level`.

```{r, eval=FALSE}
abundance_data = microbiome::abundances(pseq)
aggregate_data = microbiome::abundances(microbiome::aggregate_taxa(pseq, "Family"))
meta_data = microbiome::meta(pseq)
```

# 4. Run ANCOM on a real longitudinal dataset {.tabset}

## 4.1 Import example data

A two-week diet swap study between western (USA) and traditional (rural Africa) 
diets [@lahti2014tipping]. The dataset is available via the 
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq] 
format. In this tutorial, we consider the following fixed effects:

* Continuous covariates: "timepoint"

* Categorical covariates: "nationality"

* The group variable of interest: "group"

    + Three groups: "DI", "ED", "HE"
    
    + The reference group: "DI"
    
and the following random effects:

* A random intercept

* A random slope: "timepoint"

```{r}
data(dietswap, package = "microbiome")
print(dietswap)
```

## 4.2 Run `ancom` function using `phyloseq` data

```{r}
set.seed(123)
out = ancom(data = dietswap, tax_level = "Family", 
            p_adj_method = "holm", prv_cut = 0.10, lib_cut = 1000, 
            main_var = "group",
            adj_formula = "nationality + timepoint", 
            rand_formula = "(timepoint | subject)", 
            lme_control = lme4::lmerControl(), 
            struc_zero = TRUE, neg_lb = TRUE, alpha = 0.05, n_cl = 2)

res = out$res
```

## 4.3 Visualization for W statistics

```{r}
q_val = out$q_data
beta_val = out$beta_data
# Only consider the effect sizes with the corresponding q-value less than alpha
beta_val = beta_val * (q_val < 0.05) 
# Choose the maximum of beta's as the effect size
beta_pos = apply(abs(beta_val), 2, which.max) 
beta_max = vapply(seq_along(beta_pos), function(i) beta_val[beta_pos[i], i],
                  FUN.VALUE = double(1))
# Number of taxa included in the ALR comparisons, i.e., the taxa left after
# filtering and after the removal of taxa with structural zeros
n_taxa = nrow(out$q_data)
# Cutoff values for declaring differentially abundant taxa
cut_off = 0.7 * (n_taxa - 1)

df_fig_w = res %>%
  dplyr::mutate(beta = beta_max,
                direct = case_when(
                  detected_0.7 == TRUE & beta > 0 ~ "Positive",
                  detected_0.7 == TRUE & beta <= 0 ~ "Negative",
                  TRUE ~ "Not Significant"
                  )) %>%
  dplyr::arrange(W)
df_fig_w$taxon = factor(df_fig_w$taxon, levels = df_fig_w$taxon)
df_fig_w$W = replace(df_fig_w$W, is.infinite(df_fig_w$W), n_taxa - 1)
df_fig_w$direct = factor(df_fig_w$direct, 
                     levels = c("Negative", "Positive", "Not Significant"))

p_w = df_fig_w %>%
  ggplot(aes(x = taxon, y = W, color = direct)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(x = "Taxon", y = "W") +
  scale_color_discrete(name = NULL) + 
  geom_hline(yintercept = cut_off, linetype = "dotted", 
             color = "blue", size = 1.5) +
  geom_text(aes(x = 2, y = cut_off + 0.5, label = "W[0.7]"), 
            size = 5, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank())
p_w
```

## 4.4 Alternative input formats

The call in Section 4.2 also accepts a `TreeSummarizedExperiment` object or a
count matrix. Only the data arguments differ; all other arguments are unchanged.

For a `TreeSummarizedExperiment` object, pass `data = tse` and
`assay_name = "counts"`.

```{r, eval=FALSE}
tse = mia::makeTreeSummarizedExperimentFromPhyloseq(dietswap)
```

For a count matrix, aggregation to the analysis level is performed beforehand.
Pass `data = abundance_data`, `aggregate_data = aggregate_data`, and
`meta_data = meta_data`, and drop `tax_level`.

```{r, eval=FALSE}
abundance_data = microbiome::abundances(dietswap)
aggregate_data = microbiome::abundances(microbiome::aggregate_taxa(dietswap, "Family"))
meta_data = microbiome::meta(dietswap)
```

# Session information

```{r sessionInfo, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```

# References









