---
title: "Data Input Tutorial"
author: 
  - Menglu Liang$^1$
  - $^1$NIEHS, Research Triangle Park, NC 27709, USA
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Data Input Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, message = FALSE, warning = FALSE, comment = NA}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA,
                      fig.width = 6.25, fig.height = 5)
library(ANCOMBC)
library(tidyverse)
```

# 1. Introduction

The package requires specific format of input data. For more details, please refer to the 
[ANCOM](https://www.tandfonline.com/doi/full/10.3402/mehd.v26.27663) paper.

# 2. Installation

Download package. 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ANCOMBC")
```

Load the package. 

```{r load, eval=FALSE}
library(ANCOMBC)
```

# 3. Example Data

## 3.1 Import a phyloseq data

The HITChip Atlas dataset contains genus-level microbiota profiling with 
HITChip for 1006 western adults with no reported health complications, 
reported in [@lahti2014tipping]. The dataset is available via the 
microbiome R package [@lahti2017tools] in phyloseq [@mcmurdie2013phyloseq] 
format. 
```{r}
data(atlas1006, package = "microbiome")
dat<-data_sanity_check(atlas1006)
class(dat)
```

## 3.2 Import a tse data

We first build a TreeSummarizedExperiment Object then use it as a data example. 
```{r}
library(mia)
# microbial count table
otu_mat = matrix(sample(1:100, 100, replace = TRUE), nrow = 10, ncol = 10)
rownames(otu_mat) = paste0("taxon", 1:nrow(otu_mat))
colnames(otu_mat) = paste0("sample", 1:ncol(otu_mat))
assays = SimpleList(counts = otu_mat)

# sample metadata
smd = data.frame(group = sample(LETTERS[1:4], size = 10, replace = TRUE),row.names = paste0("sample", 1:ncol(otu_mat)),stringsAsFactors = FALSE)
smd = DataFrame(smd)
 # taxonomy table
tax_tab = matrix(sample(letters, 70, replace = TRUE),                nrow = nrow(otu_mat), ncol = 7)
rownames(tax_tab) = rownames(otu_mat)
colnames(tax_tab) = c("Kingdom", "Phylum", "Class", "Order","Family", "Genus", "Species")
# Can also contain non-taxonomic information, for instance
colnames(tax_tab) = c("G1", "G2", "G3", "G4", "G5", "G6", "G7")
tax_tab = DataFrame(tax_tab)
# create TSE
tse = TreeSummarizedExperiment(assays = assays,
                              colData = smd,
                              rowData = tax_tab)

dat<-data_sanity_check(tse)
class(dat)
```

## 3.3 import a data frame
The following R code loads the mtcars data set into our current R session. 
```{r}
data(mtcars)              # Load data frame
dat<-data_sanity_check(mtcars)
class(dat)
```



# Session information

```{r sessionInfo, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```

# References









